---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(data.table)
library(ggplot2)
library(patchwork)
library(TMB)
library(rstan)
compile("poisson_goals_model.cpp")
dyn.load(dynlib("poisson_goals_model"))
theme_set(theme_bw())
source("functions.R")
fig_dir = "figs/"
```

# FootballStanVsTMB
This repository compares Stan and TMB for modeling of football results. 

Stan is a state-of-the-art platform for statistical modeling and high-performance statistical computation. It has become the go-to tool for Bayesian modeling. The R package TMB (Template Model Builder) is a frequentist alternative that performs maximum likelihood estimation where latent variables are integrated out of the likelihood using the Laplace approximation. Both Stan and TMB use automatic differentiation (AD) for accurate evaluation of derivatives.

## Model

```{r, include=FALSE}
res = readRDS("sim_results.rds")
n_games = c(5, 10, 20, 40, 80)
dt_time = data.table(n_games = n_games * 19 * 20)
dt_time[, TMB := sapply(res, \(x) as.double(x[["tmb"]]$time, units = "secs"))]
dt_time[, Stan := sapply(res, \(x) as.double(x[["stan"]]$time, units = "secs"))]
dt_plot = melt(dt_time, id.vars = "n_games", variable.name = "Model", value.name = "Runtime")
dt_time[, relative_time := Stan / TMB]
```

## Plots

```{r}
p1 = ggplot(dt_plot, aes(n_games, Runtime, color = Model)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 0.8) + 
  scale_y_log10() + 
  xlab("Number of games") + 
  ggtitle("Runtime in seconds on log-scale")
ggsave(paste0(fig_dir, "runtime.pdf"), plot = p1, width = 12, height = 8, units = "cm")

p2 = ggplot(dt_time, aes(n_games, relative_time)) +
  geom_point(size = 2.5) + 
  geom_line(linewidth = 0.8) + 
  xlab("Number of games") + 
  ylab("Stan / TMB") + 
  ggtitle("Relative runtime")
ggsave(paste0(fig_dir, "relative_runtime.pdf"), plot = p2, width = 12, height = 8, units = "cm")

p1 + p2
```

```{r}


# Compare parameter estimates
plots = list()
for (i in c(1, 5)) {
  dt_tmb = extract_params_tmb(res[[i]]$tmb$obj)
  dt_stan = extract_params_stan(res[[i]]$stan$fit)  
  dt_tmb[, model := "TMB"]
  dt_stan[, model := "Stan"]
  keep = names(dt_tmb)
  dt_comb = rbindlist(list(dt_tmb, dt_stan[, ..keep]))
  
  plots[[paste(i)]] = 
  ggplot(dt_comb, aes(reorder(index, estimate), estimate, color = model)) + 
    facet_wrap(vars(param)) + 
    geom_point(size = 2, position=position_dodge(width=1)) + 
    geom_errorbar(aes(ymin = param_lower, ymax = param_upper),
                  width = 1, 
                  linewidth = 1.2,
                  position=position_dodge(width = 1),
                  alpha = 0.7) + 
    xlab("team") + 
    ylab("") + 
    ggtitle("Estimated parameters", subtitle = "With 95% confidence interval")  
}

plots[["1"]] / (plots[["5"]] + ggtitle("", subtitle = "")) +  plot_layout(guides = "collect") 
# ggsave(paste0(fig_dir, "param_estimates.pdf"), width = 12, height = 8, units = "cm")

```

