---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(data.table)
library(ggplot2)
library(patchwork)
library(TMB)
library(rstan)
compile("poisson_goals_model.cpp")
dyn.load(dynlib("poisson_goals_model"))
theme_set(theme_bw())
source("functions.R")
fig_dir = "figs/"
```

# FootballStanVsTMB
This repository compares Stan and TMB for modeling of football results. The model is taken from Tommy Odland's [blogg post](https://tommyodland.com/articles/2022/the-2022-norwegian-football-elite-series/). 

Stan is a state-of-the-art platform for statistical modeling and high-performance statistical computation. It has become the go-to tool for Bayesian modeling. The R package TMB (Template Model Builder) is a frequentist alternative that performs maximum likelihood estimation where latent variables are integrated out of the likelihood using the Laplace approximation. Both Stan and TMB use automatic differentiation (AD) for accurate evaluation of derivatives.

## Model

The number of goals scored by the home and away team in a match is modeled as a hierarchical model on the form: 

$$
\begin{aligned}
\text{home_goals}_i &\sim \text{poisson} \left( \theta^\text{h}_i \right) \\
\text{away_goals}_i &\sim \text{poisson} \left( \theta^\text{a}_i \right) \\
\theta^\text{h} &= \exp \left( \text{intercept} + \text{attack}_{\text{h}[i]} - \text{defence}_{\text{a}[i]} + \text{home}_{\text{h}[i]} \right) \\
\theta^\text{a} &= \exp \left( \text{intercept} +
\text{attack}_{\text{a}[i]} - \text{defence}_{\text{h}[i]} - \text{home}_{\text{h}[i]} \right) \\
\text{attack}_{\text{a}[i]} &\sim \text{normal}(0, \sigma_a^2) \\
\text{defence}_{\text{h}[i]} &\sim \text{normal}(0, \sigma_d^2) \\
\text{home}_{\text{h}[i]} &\sim \text{normal}(0, \sigma_h^2)
\end{aligned}
$$

The number of home goals is a function of a fixed intercept, the home teams attack strength minus the defense strength of the away time, and a home advantage effect. The attach, defense and home advantage effect are all random effects with normally distributed priors with unknown variance. 

For the Stan model we also have appropriate priors for the intercept and the variance of the random effects. The Stan model runs four chain of 2000 samples, where the first 1000 is used as warmup, giving us 4000 samples from the posterior distribution. 


## Run time 

The table below shows the run time for TMB and STAN for different number of observations (games). For smaller datasets TMB is about 100 times faster and for the biggest dataset of 30400 games it is about 500 times faster. In other words, a substantial speed up! This should not come as a surprise as the MCMC algorithm has to evaluate the posterior and it's gradient for each sample, while TMB uses optimization to find parameter estimates, leading to a lot fewer evaluations of the likelihood.  

```{r, echo=FALSE}
res = readRDS("sim_results.rds")
n_games = c(5, 10, 20, 40, 80)
dt_time = data.table(n_games = n_games * 19 * 20)
dt_time[, TMB := sapply(res, \(x) as.double(x[["tmb"]]$time, units = "secs"))]
dt_time[, Stan := sapply(res, \(x) as.double(x[["stan"]]$time, units = "secs"))]
dt_plot = melt(dt_time, id.vars = "n_games", variable.name = "Model", value.name = "Runtime")
dt_time[, relative_time := Stan / TMB]


knitr::kable(dt_time, digits = 1)
```

The figures below shows the run time as a function of number of games. 

```{r}
p1 = ggplot(dt_plot, aes(n_games, Runtime, color = Model)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 0.8) + 
  scale_y_log10() + 
  xlab("Number of games") + 
  ggtitle("Runtime in seconds on log-scale")
ggsave(paste0(fig_dir, "runtime.pdf"), plot = p1, width = 12, height = 8, units = "cm")

p2 = ggplot(dt_time, aes(n_games, relative_time)) +
  geom_point(size = 2.5) + 
  geom_line(linewidth = 0.8) + 
  xlab("Number of games") + 
  ylab("Stan / TMB") + 
  ggtitle("Relative runtime")
ggsave(paste0(fig_dir, "relative_runtime.pdf"), plot = p2, width = 12, height = 8, units = "cm")

p1 + p2
```


## Parameter estimates 

```{r}


# Compare parameter estimates
plots = list()
for (i in c(1, 5)) {
  dt_tmb = extract_params_tmb(res[[i]]$tmb$obj)
  dt_stan = extract_params_stan(res[[i]]$stan$fit)  
  dt_tmb[, model := "TMB"]
  dt_stan[, model := "Stan"]
  keep = names(dt_tmb)
  dt_comb = rbindlist(list(dt_tmb, dt_stan[, ..keep]))
  
  plots[[paste(i)]] = 
  ggplot(dt_comb, aes(reorder(index, estimate), estimate, color = model)) + 
    facet_wrap(vars(param)) + 
    geom_point(size = 2, position=position_dodge(width=1)) + 
    geom_errorbar(aes(ymin = param_lower, ymax = param_upper),
                  width = 1, 
                  linewidth = 1.2,
                  position=position_dodge(width = 1),
                  alpha = 0.7) + 
    xlab("team") + 
    ylab("") + 
    ggtitle("Estimated parameters", subtitle = "With 95% confidence interval")  
}

plots[["1"]] / (plots[["5"]] + ggtitle("", subtitle = "")) +  plot_layout(guides = "collect") 
# ggsave(paste0(fig_dir, "param_estimates.pdf"), width = 12, height = 8, units = "cm")

```

